Unshuffling Permutations

### 2015-09-26 submission.tex
Typos

%%
%%
%% ---- Introduction ----
%%
\section{Introduction}
\label{section:Introduction}
The {\em shuffle product}, denoted $\shuffle$, is a very famous operation on words
first defined by Eilenberg and Mac Lane~\cite{Eilenberg:MacLane:1953}.
Given three words $u$, $v_1$, and $v_2$, $u$ is said to be a \emph{shuffle}
of $v_1$ and $v_2$ if it can be formed by interleaving the letters from
$v_1$ and $v_2$ in a way that maintains the left-to-right ordering of the
letters from each word.
Besides purely combinatorial questions,
the shuffle product of words leads to the following computational ones:
\begin{enumerate}
    \item \label{item:problem_1}
    Given two words $v_1$ and $v_2$, compute the set $v_1 \shuffle v_2$.
    \item \label{item:problem_2}
    Given three words $u$, $v_1$, and $v_2$, decide if $u$ is a shuffle
    of $v_1$ and $v_2$.
    \item \label{item:problem_3}
    Given words $u$, $v_1$, \dots, $v_k$, decide if $u$ is in
    $v_1 \shuffle \dots \shuffle v_k$.
    \item \label{item:problem_4}
    Given a word $u$, decide if there is a word $v$ such that $u$ is
    in $v \shuffle v$.
\end{enumerate}
Even if these problems seem similar, they are very different in terms
of complexity. Let us now review some facts about these. In what follows,
$n$ denotes the size of $u$ and $m_i$ denotes the size of each $v_i$.
A solution to Problem~\ref{item:problem_1} can be computed in
% \begin{equation}
%     O\left((m_1 + m_2) \; \binom{m_1 + m_2}{m_1}\right)
% \end{equation}
$O\left((m_1 + m_2) \; \binom{m_1 + m_2}{m_1}\right)$
time~\cite{Spehner:TCS:1986}. An improvement and a generalization
of~Problem~\ref{item:problem_1} has been proposed
in~\cite{Allauzen:IGM:2000}, where it is proved that given words
$v_1$, \dots, $v_k$, the iterated shuffle
$v_1 \shuffle \dots \shuffle v_k$ can be computed in
% \begin{equation}
%     O\left(\binom{m_1 + \dots + m_k}{m_1, \dots, m_k}\right)
% \end{equation}
$O\left(\binom{m_1 + \dots + m_k}{m_1, \dots, m_k}\right)$
time. Problem~\ref{item:problem_2} is in \Pclass; it is indeed a
classical textbook exercise to design an efficient dynamic programming
algorithm solving it. It can be tested in $O\left(n^2 / \log(n)\right)$
time~\cite{Leeuwen:Nivat:IPL:1982}. To the best of our knowledge, the
first $O(n^2)$ time algorithm for this problem appeared
in~\cite{Mansfield:DAM:1983}. This algorithm can easily be extended to
check in polynomial-time whether or not a word is in the shuffle of any
fixed number of given words. Nevertheless, Problem~\ref{item:problem_3}
is \NPC~\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}. This
remains true even if the ground alphabet has size
$3$~\cite{Warmuth:Haussler:JCSS:1984}. Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that
Problem~\ref{item:problem_3} remains \NPC even if all the words $v_i$,
$i \in [k]$, are identical, thereby proving that, for two words $u$ and
$v$, it is \NPC to decide whether or not $u$ is in the iterated shuffle
of $v$. Again, this remains true even if the ground alphabet has size $3$.
Let us now finally focus on Problem~\ref{item:problem_4}. It is shown
in~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013} that it is \NPC to
decide if a word $u$ is a \emph{square} (w.r.t. the shuffle), that is
a word $u$ with the property that there exists a word $v$ such that $u$
is a shuffle of $v$ with itself. Hence, Problem~\ref{item:problem_4}
is \NPC.
\smallskip

This paper is intended to study a natural generalization of $\shuffle$,
denoted by $\SHUFFLE$, as a shuffle of permutations. Roughly speaking,
given three permutations $\pi$, $\sigma_1$, and $\sigma_2$, $\pi$ is
said to be a {\em shuffle} of $\sigma_1$ and $\sigma_2$ if $\pi$ (viewed
as a word) is a shuffle of two words that are order-isomorphic to
$\sigma_1$ and $\sigma_2$. Our intention in this paper is to study this
shuffle product of permutations $\SHUFFLE$ both from a combinatorial and
from a computational point of view by focusing on {\em square}
permutations, that are permutations $\pi$ being in the shuffle of a
permutation $\sigma$ with itself. The shuffle product defined and
considered in this paper is different from other existing and well-known
shuffle products on permutations. For instance, in~\cite{DHT:IJAC:2002},
the authors define the {\em shifted shuffle product}. For this product,
$\pi$ is a shuffle of $\sigma_1$ and $\sigma_2$ if $\pi$ is in the
shuffle, as words, of $\sigma_1$ and the word obtained by incrementing
all the letters of $\sigma_2$ by the size of $\sigma_1$. It is a simple
exercise to prove that, given three permutations $\pi$, $\sigma_1$, and
$\sigma_2$, deciding if $\pi$ is a in the shifted shuffle of $\sigma_1$
and $\sigma_2$ is in~\Pclass.
\smallskip

@@

Let us now provide some definitions about graphs and matching that are used in
the sequel. For a graph $G$, we denote $\VS{G}$ as the set of vertices
and $\ES{G}$ as the set of edges of $G$. Let $u = u_1 u_2 \dots u_n$ be
a word. The \emph{clique graph associated to $u$}, denoted by $K_u$, is
the complete graph on $\VS{K_u} = \{u_1, u_2, \ldots, u_n\}$. The
structure of this underlying graph is linear, \emph{i.e.}, the set of
its vertices is equipped with a natural total order $<$ defined by
$u_i < u_j$ if and only if $i < j$. For a linear graph, we write
$(u_i, u_j)$ instead of $\{u_i, u_j\}$ for an edge if it is clear from
the context that $i < j$. Recall that two edges of a graph are
\emph{independent} if they do not share a common vertex, and that a
\emph{matching} $\mathcal{M}$ in $G$ is a set of pairwise independent
edges. A matching is \emph{perfect} if it covers all the vertices of the
graph.
For an edge $e = \{u,v\}$, we shall occasionally say that $u$ and $v$
are \emph{involved} in $e$.
In case the set of vertices is equipped with a total order, a
matching $\mathcal{M}$ is said to be \emph{containment-free} if there do
not exist (independent) edges $(u_i, u_j)$ and $(u_k, u_\ell)$ in
$\mathcal{M}$ such that $u_i < u_k < u_\ell < u_j$ or
$u_k < u_i < u_j < u_\ell$
(\emph{i.e.}, $\mathcal{M}$ avoids~\MWithContainment).

@@

\begin{tabular}{c|c|c}
    Size $0$ \, & Size $2$ & Size $4$ \\ \hline
    \multirow{2}{*}{\, $\epsilon$} &
    \multirow{2}{*}{\, $12$, $21$ \,} &
    $1234$, $1243$, $1423$, $1324$, $1342$, $4132$, $3124$, $3142$,
    $3412$, $4312$, \\
    & & $2134$, $2143$, $2413$, $4213$, $2314$, $2431$, $4231$, $3241$,
    $3421$, $4321$
\end{tabular}

@@

All Sequences~\eqref{equ:sequence_square_123}, \eqref{equ:sequence_square_132},
and~\eqref{equ:sequence_square_classes} (and their subsequences obtained
by removing the $0$s) are for the time being not listed in~\cite{Slo}.

@@

\begin{lemma}
  \label{lemma:at most one edge monotone}
  Let $\pi = \pi_1 \, \sigma_1 \, \pi_2 \, \sigma_2 \, \pi_3$
  be a permutation with $|\sigma_1| \geq 2$ and $|\sigma_2| \geq 2$,
  and $\mathcal{M}$ be a containment-free perfect matching of $G_\pi$.
  The following assertions hold:
  \begin{enumerate}
    \item
    If $\sigma_1$ is increasing, $\sigma_2$ is decreasing, and
    $\sigma_1$ is above $\sigma_2$ (see Fig.~\ref{subfig:increasing before
    and above decreasing}), then $\mathcal{M}$ contains at most one
    $(\sigma_1, \sigma_2)$-edge.
    \item
    If $\sigma_1$ is decreasing, $\sigma_2$ is increasing, and $\sigma_1$
    is below $\sigma_2$ (see Fig.~\ref{subfig:decreasing before and below
    increasing}), then $\mathcal{M}$ contains at most one
    $(\sigma_1, \sigma_2)$-edge.
  \end{enumerate}
\end{lemma}

@@

  It is immediate that H can be constructed in polynomial-time in $n$ and $k$.
  % We claim that $\sigma$ occurs in $\pi$ if and only if
  % there exists a containment-free perfect matching
  % $\mathcal{M}$ in $G_\mu$ such that
  % for any two disctinct edges
  % $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  % we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.
  It can be shown (see Appendix) that
  that $\sigma$ occurs in $\pi$ if and only if
  there exists a containment-free perfect matching
  $\mathcal{M}$ in $G_\mu$ such that
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.
  \qed
\end{proof}

>>>>

%%
%%
%% ---- Introduction ----
%%
\section{Introduction}
\label{section:Introduction}
The {\em shuffle product}, denoted $\shuffle$, is a very famous operation on words
first defined by Eilenberg and Mac Lane~\cite{Eilenberg:MacLane:1953}.
Given three words $u$, $v_1$, and $v_2$, $u$ is said to be a \emph{shuffle}
of $v_1$ and $v_2$ if it can be formed by interleaving the letters from
$v_1$ and $v_2$ in a way that maintains the left-to-right ordering of the
letters from each word.
Besides purely combinatorial questions,
the shuffle product of words leads to the following algorithmic ones:
\begin{enumerate}
    \item \label{item:problem_1}
    Given two words $v_1$ and $v_2$, compute the set $v_1 \shuffle v_2$.
    \item \label{item:problem_2}
    Given three words $u$, $v_1$, and $v_2$, decide if $u$ is a shuffle
    of $v_1$ and $v_2$.
    \item \label{item:problem_3}
    Given words $u$, $v_1$, \dots, $v_k$, decide if $u$ is in
    $v_1 \shuffle \dots \shuffle v_k$.
    \item \label{item:problem_4}
    Given a word $u$, decide if there is a word $v$ such that $u$ is
    in $v \shuffle v$.
\end{enumerate}
Even if these problems seem similar, they are very different in terms
of complexity. Let us now review some facts about these. In what follows,
$n$ denotes the size of $u$ and $m_i$ denotes the size of each $v_i$.
A solution to Problem~\ref{item:problem_1} can be computed in
% \begin{equation}
%     O\left((m_1 + m_2) \; \binom{m_1 + m_2}{m_1}\right)
% \end{equation}
$O\left((m_1 + m_2) \; \binom{m_1 + m_2}{m_1}\right)$
time~\cite{Spehner:TCS:1986}. An improvement and a generalization
of~Problem~\ref{item:problem_1} has been proposed
in~\cite{Allauzen:IGM:2000}, where it is proved that given words
$v_1$, \dots, $v_k$, the iterated shuffle
$v_1 \shuffle \dots \shuffle v_k$ can be computed in
% \begin{equation}
%     O\left(\binom{m_1 + \dots + m_k}{m_1, \dots, m_k}\right)
% \end{equation}
$O\left(\binom{m_1 + \dots + m_k}{m_1, \dots, m_k}\right)$
time. Problem~\ref{item:problem_2} is in \Pclass; it is indeed a
classical textbook exercise to design an efficient dynamic programming
algorithm solving it. It can be tested in $O\left(n^2 / \log(n)\right)$
time~\cite{Leeuwen:Nivat:IPL:1982}. To the best of our knowledge, the
first $O(n^2)$ time algorithm for this problem appeared
in~\cite{Mansfield:DAM:1983}. This algorithm can easily be extended to
check in polynomial-time whether or not a word is in the shuffle of any
fixed number of given words. Nevertheless, Problem~\ref{item:problem_3}
is \NPC~\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}. This
remains true even if the ground alphabet has size
$3$~\cite{Warmuth:Haussler:JCSS:1984}. Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that
Problem~\ref{item:problem_3} remains \NPC even if all the words $v_i$,
$i \in [k]$, are identical, thereby proving that, for two words $u$ and
$v$, it is \NPC to decide whether or not $u$ is in the iterated shuffle
of $v$. Again, this remains true even if the ground alphabet has size $3$.
Let us now finally focus on Problem~\ref{item:problem_4}. It is shown
in~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013} that it is \NPC to
decide if a word $u$ is a \emph{square} (w.r.t. the shuffle), that is
a word $u$ with the property that there exists a word $v$ such that $u$
is a shuffle of $v$ with itself. Hence, Problem~\ref{item:problem_4}
is \NPC.
\smallskip

This paper is intended to study a natural generalization of $\shuffle$,
denoted by $\SHUFFLE$, as a shuffle of permutations. Roughly speaking,
given three permutations $\pi$, $\sigma_1$, and $\sigma_2$, $\pi$ is
said to be a {\em shuffle} of $\sigma_1$ and $\sigma_2$ if $\pi$ (viewed
as a word) is a shuffle of two words that are order-isomorphic to
$\sigma_1$ and $\sigma_2$. Our intention in this paper is to study this
shuffle product of permutations $\SHUFFLE$ both from a combinatorial and
from an algorithmic point of view by focusing on {\em square}
permutations, that are permutations $\pi$ being in the shuffle of a
permutation $\sigma$ with itself. The shuffle product defined and
considered in this paper is different from other existing and well-known
shuffle products on permutations. For instance, in~\cite{DHT:IJAC:2002},
the authors define the {\em shifted shuffle product}. For this product,
$\pi$ is a shuffle of $\sigma_1$ and $\sigma_2$ if $\pi$ is in the
shuffle, as words, of $\sigma_1$ and the word obtained by incrementing
all the letters of $\sigma_2$ by the size of $\sigma_1$. It is a simple
exercise to prove that, given three permutations $\pi$, $\sigma_1$, and
$\sigma_2$, deciding if $\pi$ is a in the shifted shuffle of $\sigma_1$
and $\sigma_2$ is in~\Pclass.
\smallskip

@@

Let us now provide some definitions about graphs and matchings that are used in
the sequel. For a graph $G$, we denote $\VS{G}$ as the set of vertices
and $\ES{G}$ as the set of edges of $G$. Let $u = u_1 u_2 \dots u_n$ be
a word. The \emph{clique graph associated to $u$}, denoted by $K_u$, is
the complete graph on $\VS{K_u} = \{u_1, u_2, \ldots, u_n\}$. The
structure of this underlying graph is linear, \emph{i.e.}, the set of
its vertices is equipped with a natural total order $<$ defined by
$u_i < u_j$ if and only if $i < j$. For a linear graph, we write
$(u_i, u_j)$ instead of $\{u_i, u_j\}$ for an edge if it is clear from
the context that $i < j$. Recall that two edges of a graph are
\emph{independent} if they do not share a common vertex, and that a
\emph{matching} $\mathcal{M}$ in $G$ is a set of pairwise independent
edges. A matching is \emph{perfect} if it covers all the vertices of the
graph.
For an edge $e = \{u,v\}$, we shall occasionally say that $u$ and $v$
are \emph{involved} in $e$.
In case the set of vertices is equipped with a total order, a
matching $\mathcal{M}$ is said to be \emph{containment-free} if there do
not exist (independent) edges $(u_i, u_j)$ and $(u_k, u_\ell)$ in
$\mathcal{M}$ such that $u_i < u_k < u_\ell < u_j$ or
$u_k < u_i < u_j < u_\ell$
(\emph{i.e.}, $\mathcal{M}$ avoids~\MWithContainment).

@@

\begin{tabular}{c|c|c}
    Size $0$ \, & Size $2$ & Size $4$ \\ \hline
    \multirow{2}{*}{\, $\epsilon$} &
    \multirow{2}{*}{\, $12$, $21$ \,} &
    \, $1234$, $1243$, $1423$, $1324$, $1342$, $4132$, $3124$, $3142$,
    $3412$, $4312$, \\
    & & \, $2134$, $2143$, $2413$, $4213$, $2314$, $2431$, $4231$, $3241$,
    $3421$, $4321$
\end{tabular}

@@

All Sequences~\eqref{equ:sequence_square_123}, \eqref{equ:sequence_square_132},
and~\eqref{equ:sequence_square_classes} (and their subsequences obtained
by removing the $0$'s) are for the time being not listed in~\cite{Slo}.

@@

\begin{lemma}
  \label{lemma:at most one edge monotone}
  Let $\pi = \pi_1 \, \sigma_1 \, \pi_2 \, \sigma_2 \, \pi_3$
  be a permutation with $|\sigma_1| \geq 2$ and $|\sigma_2| \geq 2$,
  and $\mathcal{M}$ be a containment-free perfect matching of $K_\pi$.
  The following assertions hold:
  \begin{enumerate}
    \item
    If $\sigma_1$ is increasing, $\sigma_2$ is decreasing, and
    $\sigma_1$ is above $\sigma_2$ (see Fig.~\ref{subfig:increasing before
    and above decreasing}), then $\mathcal{M}$ contains at most one
    $(\sigma_1, \sigma_2)$-edge.
    \item
    If $\sigma_1$ is decreasing, $\sigma_2$ is increasing, and $\sigma_1$
    is below $\sigma_2$ (see Fig.~\ref{subfig:decreasing before and below
    increasing}), then $\mathcal{M}$ contains at most one
    $(\sigma_1, \sigma_2)$-edge.
  \end{enumerate}
\end{lemma}

@@

  It is immediate that H can be constructed in polynomial-time in $n$ and $k$.
  % We claim that $\sigma$ occurs in $\pi$ if and only if
  % there exists a containment-free perfect matching
  % $\mathcal{M}$ in $G_\mu$ such that
  % for any two disctinct edges
  % $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  % we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.
  It can be shown (see Appendix) that
  $\sigma$ occurs in $\pi$ if and only if
  there exists a containment-free perfect matching
  $\mathcal{M}$ in $K_\mu$ such that
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.
  \qed
\end{proof}
